### 2022-07-27 paper/localpsf.tex
edited paper during discussion with Tucker

We present a fast \emph{matrix free} method for approximating operators, $\Aop:L^2(\Omega) \rightarrow L^2(\Omega)'$, that have locally supported non-negative integral kernels. Here $\Omega \subset \mathbb{R}^\gdim$ is a bounded domain, $\gdim$ is the spatial dimension, and $L^2(\Omega)'$ is the space of continuous linear functionals on $L^2(\Omega)$ (i.e., the topological dual space). Such operators arise as Schur complements in Schur complements methods for solving partial differential equations (PDEs) \cite{le1997non,saad1999distributed,TOBYPAPER}, Poincare-Steklov operators in domain decomposition methods (e.g., Dirichlet-to-Neumann maps) \cite{DTNMAP}, covariance operators in spatial statistics \cite{GeogaEtAl20,ChenStein21,LindgrenRueLindstrom11}, blurring operators in imaging \cite{DenisEtAl11,NagyOleary98}, and Hessians in distributed parameter PDE-constrained optimization and inverse problems \cite{AlgerEtAl19,IsaacEtAl15}.

By ``matrix free,'' we mean that we have a black-box computational procedure through which we may apply $\Aop$ and its transpose\footnote{Recall that the transpose $\Aop^T:L^2(\Omega)\rightarrow L^2(\Omega)'$ is the unique operator satisfying $\left(\Aop u\right)(w) = \left(\Aop w\right)(u)$ for all $u,w \in L^2(\Omega)$, where the linear functional $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, the scalar $\left(\Aop u\right)(w)$ is the result of applying that linear functional to $w \in L^2(\Omega)$, and similar for the analogous operations with $\Aop^T$. }, $\Aop^T$, to arbitrary functions,
\begin{equation}
\label{eq:eval_maps}
u \mapsto\Aop u \qquad \text{and} \qquad w \mapsto \Aop^T w,
\end{equation} 
but we do not have any further access to $\Aop$. In particular, we cannot easily access entries of $\Aop$'s integral kernel. In applications, evaluating the maps in \eqref{eq:eval_maps} may involve iteratively solving a large linear system, timestepping, or performing some other costly computational procedure.

Our motivation is Hessian approximation in distributed parameter inverse problems governed by partial differential equations (PDEs). That is, inverse problems in which one seeks to reconstruct an unknown parameter field from noisy observations of a state variable that depends on the parameter implicitly through the solution of a PDE.
In the deterministic approach to the inverse problem, one solves a regularized minimization problem to find the parameter that best fits the data. The objective function in the minimization problem is defined implicitly, so information about the Hessian of the objective function is only accessible via application of the Hessian to vectors \cite{HESSIANACTION}. 
The most popular matrix-free Hessian approximation methods are based on low rank approximation \cite{BuiEtAl13,CuiEtAl14,FlathEtAl11,PetraEtAl14,SpantiniEtAl15} because there are efficient methods for constructing low rank approximations of a matrix using only matrix-vector products \cite{Cheng05,HalkoMartinssonTropp11}. 
However, low rank approximation methods suffer from a ``data predicament''---if the data are highly informative about the unknown parameter, then the numerical rank of the data misfit term in the Hessian is large, so a large number of operator applies are required to form an accurate ``low rank'' approximation \cite{Alger19,AlgerEtAl17}. High-rank approximation methods are needed.

@@

The central challenge we face is that existing scalable high-rank approximation methods typically require the ability to rapidly evaluate arbitrary entries of $\Aop$'s integral kernel, and we cannot do this because $\Aop$ is only accessible matrix-free. We address this challenge using \emph{impulse response interpolation}. The impulse response, $\phi_x$, associated with a point $x$ is the Riesz representation\footnote{Recall that the Riesz representative of a functional $\rho \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\rho^* \in L^2(\Omega)$ such that $\rho(w) = \left(\rho^*,w\right)_{L^2(\Omega)}$ for all $w \in L^2(\Omega)$.} of the linear functional that results from applying $\Aop$ to a delta distribution (i.e., point source, impulse) centered at $x$. We compute batches of impulse responses by applying $\Aop$ to weighted sums of delta distributions associated with batches of points scattered throughout the domain (see Figure \ref{fig:batches_intro}). Then we interpolate translated versions of these impulse responses to approximate arbitrary entries of the operator's integral kernel. 
%This allows fast evaluation of approximate matrix entries, which are needed by high-rank approximation methods, in particular, classical H-matrix construction methods.

Picking the batches of points requires us to estimate the supports of the $\phi_x$, \emph{before} we compute them. The idea of estimating the supports of the functions $\phi_x$ a-priori was inspired by techniques from resolution analysis in seismic imaging, in which the width of $\phi_x$ is estimated to be the local autocorrelation length of the function $\Aop^T \zeta$ near $x$, where $\zeta$ is a random noise funtion  \cite{FichtnerLeeuwen15,TrampertFichtnerRitsema13}. Rather than probing $\Aop^T$ with random noise functions, we use polynomial functions (see Section \ref{sec:intromoments}). Our method estimates the support of $\phi_x$ more accurately and reliably than random noise probing at the cost of the additional constraint that $\Aop$ has a positive integral kernel.

@@

The ability to rapidly evaluate approximations of arbitrary entries of $\Aop$'s integral kernel allows us to form a hierarchical matrix (H-matrix) \cite{Hackbusch99,} approximation of a discretized version of $\Aop$. H-matrices are a matrix format in which the rows and columns of the matrix are re-ordered, then the matrix is recursively subdivided into blocks, in such a way that many off-diagonal blocks are low rank, even though the matrix as a whole may be high-rank. 
%(see Figure \ref{fig:hmatrix_intro} and Appendix \ref{app:h_matrix}).
H-matrix methods permit us to perform matrix-vector products cheaply. Furthermore, one may use H-matrix methods to perform useful linear algebra operations involving the H-matrix that cannot easily be done using the original operator. These operations include matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion. 
%These H-matrix methods are fast and scale well to large problems. 
The work and memory required to perform these operations for an $\fedim \times \fedim$ H-matrix is scales as $O(\fedim \log(\fedim)^a)$, where $a \in \{0,1,2,3\}$. That is, the cost is \emph{nearly-linear} in $N$. The exact cost depends on the type of H-matrix used, the operation being performed, and the rank of the off-diagonal blocks; for more details see \cite{GrasedyckHackbusch03}. The ability to perform these operations permits, for example, fast solution of Newton linear systems in PDE-constrained optimization, fast sampling of ill-conditioned posterior distribution in Bayesian inverse problems, and construction of high rank surrogate models that can be used for uncertainty quantification.

@@

Let $\febasis_1, \febasis_2, \dots, \febasis_\fedim$ be a set of continuous Galerkin finite element basis functions used to discretize the problem on a mesh with mesh size parameter $h$, let $V_h := \Span\left(\febasis_1, \febasis_2, \dots, \febasis_\fedim\right)$ be the corresponding finite element space under the $L^2$ inner product, and let $p_i \in \mathbb{R}^\gdim$, $i=1,\dots, \fedim$ be the Lagrange nodes associated with the functions $\febasis_i$. We assume that the finite element basis satisfies the Kronecker property, i.e., $\febasis_i(p_i)=1$ and $\febasis_i(p_j)=0$ if $i\neq j$. 
For $u_h \in V_h$ we write $\bm{u} \in \mathbb{R}^\nsamplepts_\massmatrix$ to denote the coefficient vector for $u_h$ with respect to the finite element basis, i.e.,
\begin{equation}
\label{eq:fem_coeff_basis}
u_h(x) = \sum_{i=1}^\fedim \bm{u}_i \febasis_i(x).
\end{equation}
Linear functionals $\rho_h \in V_h'$ have coefficient dual vectors $\boldsymbol{\rho}\in \mathbb{R}^\nsamplepts_{\massmatrix^{-1}}$, with entries 
\begin{equation*}
\boldsymbol{\rho}_i = \rho_h(\febasis_i), \qquad i=1,\dots,\nsamplepts. 
\end{equation*}
Here $\massmatrix \in \mathbb{R}^{\fedim \times \fedim}$ denotes the sparse finite element mass matrix which has entries
\begin{equation*}
\massmatrix_{ij}=\int_\Omega \febasis_i(x) \febasis_j(x) dx, \qquad i,j=1,\dots,\fedim,
\end{equation*}
the space $\mathbb{R}^\fedim_\massmatrix$ is $\mathbb{R}^\fedim$ with the inner product $(\mathbf{u},\mathbf{w})_\massmatrix := \mathbf{u}^T \massmatrix \mathbf{w}$, and $\mathbb{R}^\fedim_{\massmatrix^{-1}}$ is the analogous space with $\massmatrix^{-1}$ replacing $\massmatrix$. Direct calculation shows that $\mathbb{R}^\fedim_\massmatrix$ and $\mathbb{R}^\fedim_{\massmatrix^{-1}}$ are isomorphic to $V_h$ and $V_h'$ as Hilbert spaces, respectively.

@@

The operator $A_h:V_h \rightarrow V_h'$ can be written in integral kernel form as
\begin{equation}
(A_h u_h)(w_h) := \int_\Omega \int_\Omega w_h(y) \Aker_h(y,x) u_h(x) dx dy
\end{equation}
for some kernel $\Aker_h$ which we do not know, and which may differ from $\Aker$ due to discretization error. Since the functions in $V_h$ are continuous at $x$, the delta distribution $\delta_x$ is a continuous linear functional on $V_h$, which has a discrete dual vector $\boldsymbol{\delta}_x \in \mathbb{R}^\fedim_{\massmatrix^{-1}}$ with entries $\left(\boldsymbol{\delta}_x\right)_i = \febasis_i(x)$ for $i=1,\dots,\fedim$. Additionally, it is straightforward to verify that the Riesz representation, $\rho_h^* \in V_h$, of a functional $\rho \in V_h'$ has coefficient vector
\begin{equation*}
\boldsymbol{\rho}^* = \massmatrix^{-1} \boldsymbol{\rho}.
\end{equation*}
Therefore, the formula for the impulse response from \eqref{eq:impulse_response_delta_action} becomes 
\begin{equation}
\label{eq:discrete_kernel}
\boldsymbol{\impulseresponse}_x = \left(A_h \delta_x^*\right)^* =  \massmatrix^{-1}\mathbf{A} \massmatrix^{-1} \boldsymbol{\delta}_x,
\end{equation}
and the $(y,x)$ kernel entry of $\Aker_h$ may be written as
\begin{equation}
\label{eq:approx_kernel_entry_81}
	\Aker_h(y,x) = \boldsymbol{\delta}_y^T \boldsymbol{\impulseresponse}_x = \boldsymbol{\delta}_y^T \massmatrix^{-1}\mathbf{A} \massmatrix^{-1} \boldsymbol{\delta}_x.
\end{equation}
Now define $\mathbf{\Aker} \in \mathbb{R}^{\fedim \times \fedim}$ to be the following dense matrix of kernel entries evaluated at all pairs of Lagrange nodes:
\begin{equation}
\label{eq:Akerpcmat_entries}
\mathbf{\Aker}_{ij} := \Aker_h(p_i, p_j).
\end{equation}
Because of the Kronecker property of the finite element basis, we have $\boldsymbol{\delta}_{p_i} = \mathbf{e}_i$. Thus from \eqref{eq:approx_kernel_entry_81} we have
$
	\Aker_h(p_i,p_j) = \left(\massmatrix^{-1}\mathbf{A} \massmatrix^{-1}\right)_{ij},
$
which implies
\begin{equation}
\label{eq:boldA}
	\mathbf{A} = \massmatrix \mathbf{\Aker} \massmatrix.
\end{equation}
Broadly, our method will construct an H-matrix approximation of $\mathbf{A}$ by forming H-matrix approximations of $\boldsymbol{\Aker}$ and $\massmatrix$ (or a lumped mass version of $\massmatrix$), then multiplying these matrices with H-matrix methods per \eqref{eq:boldA}. The central challenge is that classical H-matrix construction methods require access to arbitrary matrix entries $\mathbf{\Aker}_{ij}$, but these matrix entries are not easily accessible. The bulk of our method is therefore dedicated to forming approximations of these matrix entries that can be evaluated rapidly.

@@

Approximation \eqref{eq:translate1} says that $\phi_x$ can be approximated by $\phi_{x'}$ when $x$ and $x'$ are close. Operators satisfying \eqref{eq:translate1} can be well-approximated via low-rank CUR approximation \cite{CUR,NYSTROM}. However, the required rank in the low rank approximation can be large, which makes algorithms based on \eqref{eq:translate1} expensive.

@@

The ability to rapidly evaluate approximate kernel entries $\Aker(y,x)$ allows us to construct an H-matrix approximation, $\boldsymbol{\Aker}_H \approx \mathbf{\Aker}$, using the conventional adaptive cross H-matrix construction method (Section \ref{sec:h_matrix_construction_short}). In this method, one forms low-rank approximations of off-diagonal blocks of the matrix by sampling rows and columns of those blocks \cite{HACA}. We then use H-matrix methods to convert $\boldsymbol{\Aker}_H$ into an H-matrix approximation $\mathbf{A}_H \approx \mathbf{A}$.

@@

We start with a finite set of candidate points $\candidatepts$ and build $\pointbatch_k$ incrementally with points selected from $\candidatepts$. For simplicity of explanation, here $\pointbatch_k$ and $\candidatepts$ are mutable sets that we add points to and remove points from. First we initialize $\pointbatch_k$ as an empty set. Then we select the candidate point $\candidatepoint_i \in \candidatepts$ that is the farthest away from all points in previous sample point batches $S_1 \cup S_2 \cup \dots \cup S_{k-1}$. Candidate points for the first batch $S_1$ are chosen randomly from $\candidatepts$.
Once $\candidatepoint_i$ is selected, we remove $\candidatepoint_i$ from $\candidatepts$. Then we perform the following two checks:
\begin{enumerate}
	\item We check whether $\candidatepoint_i$ is sufficiently far from all of the previously chosen points in the current batch, in the sense that $E_{\candidatepoint_i} \cap E_{\candidatepoint_j} = \{\}$ for all $\candidatepoint_j \in \pointbatch_k$.
	\item We make sure that $\spatialvol(\candidatepoint_i)$ is not too small, by checking whether $\spatialvol(\candidatepoint_i) > \epsilon \spatialvol_\text{max}$. Here $\spatialvol_\text{max}$ is the largest value of $\spatialvol(\candidatepoint_j)$ over all points $q$ in the initial set of candidate points, and $\tau$ is a small threshold parameter (we use $\epsilon=10^{-5}$).
\end{enumerate}
If $\candidatepoint_i$ passes both these checks (i.e., if $\candidatepoint_i$ is sufficiently far from other points in the batch, and $\spatialvol(\candidatepoint_i)$ is not too small) then we add $\candidatepoint_i$ to $\pointbatch_k$. Otherwise we discard $\candidatepoint_i$. This process repeats until there are no more points in $\candidatepts$.  
%This is detailed in Algorithm \ref{alg:point_choice}. 
We check whether $E_{\candidatepoint_i} \cap E_{\candidatepoint_j} = \{\}$ using the ellipsoid intersection test described in Appendix \ref{sec:fast_ellipsoid_intersection_test}. We repeat the process to construct several batches of points $\pointbatch_1, \pointbatch_2, \dots, \pointbatch_{\nbatch}$. In our implementation, for each batch, $\candidatepts$ is initialized as the set of all Lagrange nodes for the finite element basis functions used to discretize the problem, except for points in previously chosen batches.

@@

Given $(y,x)\in \Omega \times \Omega$, let
\begin{equation*}
	z_i := y - \spatialmean(x) + \spatialmean(x_i), \qquad i=1,\dots,\numnbr,
\end{equation*}
where $\{x_i\}_{i=1}^{\numnbr}$ are the $\numnbr$ nearest sample points to $x$, excluding sample points $x_i$ for which $z_i \notin \Omega$. Here $\numnbr$ is a small user defined parameter; e.g., $\numnbr=10$. To find the $\numnbr$ nearest sample points to $x$, we query a precomputed k-d tree \cite{Bentley75} of all sample points.  We check whether $z_i \in \Omega$ by querying a precomputed axis aligned bounding box tree (AABB tree) \cite{Ericson04} of the mesh cells used to discretize the problem. Further, let
\begin{equation}
\label{eq:fxyxp}
f_i := \frac{\spatialvol(x)}{\spatialvol(x_i)}\phi_{x_i}\left(z_i\right), \qquad i=1,\dots,\numnbr,
\end{equation}
Note that $\phi_{x_i}\left(z_i\right)$ is well-defined because $z_i \in \Omega$, and $\frac{\spatialvol(x)}{\spatialvol(x_i)}$ is well-defined because our sample point picking procedure in Section \ref{sec:sample_point_selection} ensures that $\spatialvol(x_i)>0$. We approximate $\Aker(y,x)$ by interpolating the (point,value) pairs
\begin{equation}
\left(x_i, f_i\right), \qquad i=1,\dots,\numnbr,
\end{equation}
at the point $x$. The idea is that $\Aker(y,x) = f_i$ per the discussion in Section \ref{sec:local_mean_displacement_invariance}, so we approximate $\Aker(y,x)$ by interpolating the values $f_i$.

@@

\begin{figure}[H]
	\begin{subfigure}{0.5\textwidth}
		\includegraphics[scale=0.23]{meshHeight_view2_edges.png}
		\caption{Bird's eye view}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\includegraphics[scale=0.22]{meshHeight_view0_edges_withAxes.png}
		\caption{Top-down view}
	\end{subfigure}
	\caption{The ice sheet, discretized by a mesh of tetrahedra. Color indicates the height of the base of the ice sheet (i.e., the mountain topography). The radius of the domain is $10$ kilometers, the maximum height of the mountain is $2.1$ kilometers, and the average thickness of the ice sheet is 250 meters.}
	\label{fig:ice_mountain_mesh}
\end{figure}

>>>>

We present a fast \emph{matrix-free} method for approximating operators, $\Aop:L^2(\Omega) \rightarrow L^2(\Omega)'$, that have locally supported non-negative integral kernels. Here $\Omega \subset \mathbb{R}^\gdim$ is a bounded domain, $\gdim$ is the spatial dimension, and $L^2(\Omega)'$ is the space of continuous linear functionals on $L^2(\Omega)$ (i.e., the topological dual space). Such operators arise as Schur complements in Schur complements methods for solving partial differential equations (PDEs) \cite{le1997non,saad1999distributed,TOBYPAPER}, Poincare-Steklov operators in domain decomposition methods (e.g., Dirichlet-to-Neumann maps) \cite{DTNMAP}, covariance operators in spatial statistics \cite{GeogaEtAl20,ChenStein21,LindgrenRueLindstrom11}, blurring operators in imaging \cite{DenisEtAl11,NagyOleary98}, and Hessians in distributed parameter PDE-constrained optimization and inverse problems \cite{AlgerEtAl19,IsaacEtAl15}.

By ``matrix-free,'' we mean that we have a black-box computational procedure through which we may apply $\Aop$ and its transpose\footnote{Recall that the transpose $\Aop^T:L^2(\Omega)\rightarrow L^2(\Omega)'$ is the unique operator satisfying $\left(\Aop u\right)(w) = \left(\Aop w\right)(u)$ for all $u,w \in L^2(\Omega)$, where the linear functional $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, the scalar $\left(\Aop u\right)(w)$ is the result of applying that linear functional to $w \in L^2(\Omega)$, and similar for the analogous operations with $\Aop^T$. }, $\Aop^T$, to arbitrary functions,
\begin{equation}
\label{eq:eval_maps}
u \mapsto\Aop u \qquad \text{and} \qquad w \mapsto \Aop^T w,
\end{equation} 
but we do not have any further access to $\Aop$. In particular, we cannot easily access entries of $\Aop$'s integral kernel. In applications, evaluating the maps in \eqref{eq:eval_maps} may involve iteratively solving a large linear system, timestepping, or performing some other costly computational procedure.

Our motivation is Hessian approximation in distributed parameter inverse problems governed by partial differential equations (PDEs). That is, inverse problems in which one seeks to reconstruct an unknown parameter field from noisy observations of a state variable that depends on the parameter implicitly through the solution of a PDE.
In the deterministic approach to the inverse problem, one solves a regularized minimization problem to find the parameter that best fits the data. The objective function in the minimization problem is defined implicitly, so information about the Hessian of the objective function is only accessible via application of the Hessian to vectors \cite{HESSIANACTION}. 
The most popular matrix-free Hessian approximation methods are based on low rank approximation \cite{BuiEtAl13,CuiEtAl14,FlathEtAl11,PetraEtAl14,SpantiniEtAl15} because there are efficient methods for constructing low rank approximations of a matrix using only matrix-vector products \cite{Cheng05,HalkoMartinssonTropp11}. 
However, low rank approximation methods suffer from a ``data predicament''---if the data are highly informative about the unknown parameter, then the numerical rank of the data misfit term in the Hessian is large, so a large number of operator applies are required to form an accurate ``low rank'' approximation \cite{Alger19,AlgerEtAl17}. High rank approximation methods are needed.

@@

The central challenge we face is that existing scalable high rank approximation methods typically require the ability to rapidly evaluate arbitrary entries of $\Aop$'s integral kernel, and we cannot do this because $\Aop$ is only accessible matrix-free. We address this challenge using \emph{impulse response interpolation}. The impulse response, $\phi_x$, associated with a point $x$ is the Riesz representation\footnote{Recall that the Riesz representative of a functional $\rho \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\rho^* \in L^2(\Omega)$ such that $\rho(w) = \left(\rho^*,w\right)_{L^2(\Omega)}$ for all $w \in L^2(\Omega)$.} of the linear functional that results from applying $\Aop$ to a delta distribution (i.e., point source, impulse) centered at $x$. We compute batches of impulse responses by applying $\Aop$ to weighted sums of delta distributions associated with batches of points scattered throughout the domain (see Figure \ref{fig:batches_intro}). Then we interpolate translated versions of these impulse responses to approximate arbitrary entries of the operator's integral kernel. 
%This allows fast evaluation of approximate matrix entries, which are needed by high-rank approximation methods, in particular, classical H-matrix construction methods.

Picking the batches of points requires us to estimate the supports of the $\phi_x$, \emph{before} we compute them. The idea of estimating the supports of the functions $\phi_x$ a-priori was inspired by techniques from resolution analysis in seismic imaging. There, $\Aop^T$ is applied to a random noise function, and the width of $\phi_x$ is estimated to be the autocorrelation length of the resultant function near $x$  \cite{FichtnerLeeuwen15,TrampertFichtnerRitsema13}. Rather than probing $\Aop^T$ with random noise functions, we use polynomial functions (see Section \ref{sec:intromoments}). Our method estimates the support of $\phi_x$ more accurately and reliably than random noise probing at the cost of the additional constraint that $\Aop$ has a positive integral kernel.

@@

The ability to rapidly approximate entries of $\Aop$'s integral kernel allows us to form a hierarchical matrix (H-matrix) \cite{Hackbusch99,} approximation of a discretized version of $\Aop$. H-matrices are a matrix format in which the rows and columns of the matrix are re-ordered, then the matrix is recursively subdivided into blocks, in such a way that many off-diagonal blocks are low rank, even though the matrix as a whole may be high rank. 
%(see Figure \ref{fig:hmatrix_intro} and Appendix \ref{app:h_matrix}).
H-matrix methods permit us to perform matrix-vector products cheaply, and perform other useful linear algebra operations that cannot be done easily using the original operator. These operations include matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion. 
%These H-matrix methods are fast and scale well to large problems. 
The work and memory required to perform these operations for an $\fedim \times \fedim$ H-matrix scales 
%as $O(\fedim \log(\fedim)^a)$, where $a \in \{0,1,2,3\}$. That is, the cost is \emph{nearly-linear}
\emph{nearly-linearly} in $N$ (i.e., $o(N^{1+\epsilon})$ for any $\epsilon > 0$). The exact cost depends on the type of H-matrix used, the operation being performed, and the rank of the off-diagonal blocks; for more details see \cite{GrasedyckHackbusch03}. The ability to perform these operations permits, for example, fast solution of Newton linear systems in PDE-constrained optimization, fast sampling of ill-conditioned posterior distribution in Bayesian inverse problems, and construction of high rank surrogate models that can be used for uncertainty quantification.

@@

Let $\febasis_1, \febasis_2, \dots, \febasis_\fedim$ be a set of continuous Galerkin finite element basis functions used to discretize the problem on a mesh with mesh size parameter $h$, let $V_h := \Span\left(\febasis_1, \febasis_2, \dots, \febasis_\fedim\right)$ be the corresponding finite element space under the $L^2$ inner product, and let $\fepoint_i \in \mathbb{R}^\gdim$, $i=1,\dots, \fedim$ be the Lagrange nodes associated with the functions $\febasis_i$. We assume that the finite element basis satisfies the Kronecker property, i.e., $\febasis_i(\fepoint_i)=1$ and $\febasis_i(\fepoint_j)=0$ if $i\neq j$. 
For $u_h \in V_h$ we write $\bm{u} \in \mathbb{R}^\nsamplepts_\massmatrix$ to denote the coefficient vector for $u_h$ with respect to the finite element basis, i.e.,
\begin{equation}
\label{eq:fem_coeff_basis}
u_h(x) = \sum_{i=1}^\fedim \bm{u}_i \febasis_i(x).
\end{equation}
Linear functionals $\rho_h \in V_h'$ have coefficient dual vectors $\boldsymbol{\rho}\in \mathbb{R}^\nsamplepts_{\massmatrix^{-1}}$, with entries 
\begin{equation*}
\boldsymbol{\rho}_i = \rho_h(\febasis_i), \qquad i=1,\dots,\nsamplepts. 
\end{equation*}
Here $\massmatrix \in \mathbb{R}^{\fedim \times \fedim}$ denotes the sparse finite element mass matrix which has entries
\begin{equation*}
\massmatrix_{ij}=\int_\Omega \febasis_i(x) \febasis_j(x) dx, \qquad i,j=1,\dots,\fedim,
\end{equation*}
the space $\mathbb{R}^\fedim_\massmatrix$ is $\mathbb{R}^\fedim$ with the inner product $(\mathbf{u},\mathbf{w})_\massmatrix := \mathbf{u}^T \massmatrix \mathbf{w}$, and $\mathbb{R}^\fedim_{\massmatrix^{-1}}$ is the analogous space with $\massmatrix^{-1}$ replacing $\massmatrix$. Direct calculation shows that $\mathbb{R}^\fedim_\massmatrix$ and $\mathbb{R}^\fedim_{\massmatrix^{-1}}$ are isomorphic to $V_h$ and $V_h'$ as Hilbert spaces, respectively.

@@

The operator $A_h:V_h \rightarrow V_h'$ can be written in integral kernel form as
\begin{equation}
(A_h u_h)(w_h) := \int_\Omega \int_\Omega w_h(y) \Aker_h(y,x) u_h(x) dx dy
\end{equation}
for some kernel $\Aker_h$ which we do not know, and which may differ from $\Aker$ due to discretization error. Since the functions in $V_h$ are continuous at $x$, the delta distribution $\delta_x$ is a continuous linear functional on $V_h$, which has a discrete dual vector $\boldsymbol{\delta}_x \in \mathbb{R}^\fedim_{\massmatrix^{-1}}$ with entries $\left(\boldsymbol{\delta}_x\right)_i = \febasis_i(x)$ for $i=1,\dots,\fedim$. Additionally, it is straightforward to verify that the Riesz representation, $\rho_h^* \in V_h$, of a functional $\rho \in V_h'$ has coefficient vector
\begin{equation*}
\boldsymbol{\rho}^* = \massmatrix^{-1} \boldsymbol{\rho}.
\end{equation*}
Therefore, the formula for the impulse response from \eqref{eq:impulse_response_delta_action} becomes 
\begin{equation}
\label{eq:discrete_kernel}
\boldsymbol{\impulseresponse}_x = \left(A_h \delta_x^*\right)^* =  \massmatrix^{-1}\mathbf{A} \massmatrix^{-1} \boldsymbol{\delta}_x,
\end{equation}
and the $(y,x)$ kernel entry of $\Aker_h$ may be written as
\begin{equation}
\label{eq:approx_kernel_entry_81}
	\Aker_h(y,x) = \boldsymbol{\delta}_y^T \boldsymbol{\impulseresponse}_x = \boldsymbol{\delta}_y^T \massmatrix^{-1}\mathbf{A} \massmatrix^{-1} \boldsymbol{\delta}_x.
\end{equation}
Now define $\mathbf{\Aker} \in \mathbb{R}^{\fedim \times \fedim}$ to be the following dense matrix of kernel entries evaluated at all pairs of Lagrange nodes:
\begin{equation}
\label{eq:Akerpcmat_entries}
\mathbf{\Aker}_{ij} := \Aker_h(\fepoint_i, \fepoint_j).
\end{equation}
Because of the Kronecker property of the finite element basis, we have $\boldsymbol{\delta}_{\fepoint_i} = \mathbf{e}_i$. Thus from \eqref{eq:approx_kernel_entry_81} we have
$
	\Aker_h(\fepoint_i,\fepoint_j) = \left(\massmatrix^{-1}\mathbf{A} \massmatrix^{-1}\right)_{ij},
$
which implies
\begin{equation}
\label{eq:boldA}
	\mathbf{A} = \massmatrix \mathbf{\Aker} \massmatrix.
\end{equation}
Broadly, our method will construct an H-matrix approximation of $\mathbf{A}$ by forming H-matrix approximations of $\boldsymbol{\Aker}$ and $\massmatrix$ (or a lumped mass version of $\massmatrix$), then multiplying these matrices with H-matrix methods per \eqref{eq:boldA}. The central challenge is that classical H-matrix construction methods require access to arbitrary matrix entries $\mathbf{\Aker}_{ij}$, but these matrix entries are not easily accessible. The bulk of our method is therefore dedicated to forming approximations of these matrix entries that can be evaluated rapidly.

@@

Approximation \eqref{eq:translate1} says that $\phi_x$ can be approximated by $\phi_{x'}$ when $x$ and $x'$ are close. Operators satisfying \eqref{eq:translate1} can be well-approximated via low rank CUR approximation \cite{CUR,NYSTROM}. However, the required rank in the low rank approximation can be large, which makes algorithms based on \eqref{eq:translate1} expensive.

@@

The ability to rapidly evaluate approximate kernel entries $\Aker(y,x)$ allows us to construct an H-matrix approximation, $\boldsymbol{\Aker}_H \approx \mathbf{\Aker}$, using the conventional adaptive cross H-matrix construction method (Section \ref{sec:h_matrix_construction_short}). In this method, one forms low rank approximations of off-diagonal blocks of the matrix by sampling rows and columns of those blocks \cite{HACA}. We then use H-matrix methods to convert $\boldsymbol{\Aker}_H$ into an H-matrix approximation $\mathbf{A}_H \approx \mathbf{A}$.

@@

We start with a finite set of candidate points $\candidatepts$ and build $\pointbatch_k$ incrementally with points selected from $\candidatepts$. For simplicity of explanation, here $\pointbatch_k$ and $\candidatepts$ are mutable sets that we add points to and remove points from. First we initialize $\pointbatch_k$ as an empty set. Then we select the candidate point $\candidatepoint_i \in \candidatepts$ that is the farthest away from all points in previous sample point batches $S_1 \cup S_2 \cup \dots \cup S_{k-1}$. Candidate points for the first batch $S_1$ are chosen randomly from $\candidatepts$.
Once $\candidatepoint_i$ is selected, we remove $\candidatepoint_i$ from $\candidatepts$. Then we perform the following two checks:
\begin{enumerate}
	\item We check whether $\candidatepoint_i$ is sufficiently far from all of the previously chosen points in the current batch, in the sense that $E_{\candidatepoint_i} \cap E_{\candidatepoint_j} = \{\}$ for all $\candidatepoint_j \in \pointbatch_k$.
	\item We make sure that $\spatialvol(\candidatepoint_i)$ is not too small, by checking whether $\spatialvol(\candidatepoint_i) > \epsilon \spatialvol_\text{max}$. Here $\spatialvol_\text{max}$ is the largest value of $\spatialvol(\candidatepoint_j)$ over all points $q$ in the initial set of candidate points, and $\epsilon$ is a small threshold parameter (we use $\epsilon=10^{-5}$).
\end{enumerate}
If $\candidatepoint_i$ passes both these checks (i.e., if $\candidatepoint_i$ is sufficiently far from other points in the batch, and $\spatialvol(\candidatepoint_i)$ is not too small) then we add $\candidatepoint_i$ to $\pointbatch_k$. Otherwise we discard $\candidatepoint_i$. This process repeats until there are no more points in $\candidatepts$.  
%This is detailed in Algorithm \ref{alg:point_choice}. 
We check whether $E_{\candidatepoint_i} \cap E_{\candidatepoint_j} = \{\}$ using the ellipsoid intersection test described in Appendix \ref{sec:fast_ellipsoid_intersection_test}. We repeat the process to construct several batches of points $\pointbatch_1, \pointbatch_2, \dots, \pointbatch_{\nbatch}$. In our implementation, for each batch, $\candidatepts$ is initialized as the set of all Lagrange nodes for the finite element basis functions used to discretize the problem, except for points in previously chosen batches.

@@

Given $(y,x)\in \Omega \times \Omega$, let
\begin{equation*}
	z_i := y - \spatialmean(x) + \spatialmean(x_i), \qquad i=1,\dots,\numnbr,
\end{equation*}
where $\{x_i\}_{i=1}^{\numnbr}$ are the $\numnbr$ nearest sample points to $x$, excluding sample points $x_i$ for which $z_i \notin \Omega$. Here $\numnbr$ is a small user defined parameter; e.g., $\numnbr=10$. To find the $\numnbr$ nearest sample points to $x$, we query a precomputed kd-tree \cite{Bentley75} of all sample points.  We check whether $z_i \in \Omega$ by querying a precomputed axis aligned bounding box tree (AABB tree) \cite{Ericson04} of the mesh cells used to discretize the problem. Further, let
\begin{equation}
\label{eq:fxyxp}
f_i := \frac{\spatialvol(x)}{\spatialvol(x_i)}\phi_{x_i}\left(z_i\right), \qquad i=1,\dots,\numnbr,
\end{equation}
Note that $\phi_{x_i}\left(z_i\right)$ is well-defined because $z_i \in \Omega$, and $\frac{\spatialvol(x)}{\spatialvol(x_i)}$ is well-defined because our sample point picking procedure in Section \ref{sec:sample_point_selection} ensures that $\spatialvol(x_i)>0$. We approximate $\Aker(y,x)$ by interpolating the (point,value) pairs
\begin{equation}
\left(x_i, f_i\right), \qquad i=1,\dots,\numnbr,
\end{equation}
at the point $x$. The idea is that $\Aker(y,x) = f_i$ per the discussion in Section \ref{sec:local_mean_displacement_invariance}, so we approximate $\Aker(y,x)$ by interpolating the values $f_i$.

@@

\begin{figure}[H]
	\begin{subfigure}{0.5\textwidth}
		\includegraphics[scale=0.23]{meshHeight_view2_edges.png}
		\caption{Bird's eye view}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\includegraphics[scale=0.22]{meshHeight_view0_edges_withAxes.png}
		\caption{Top-down view}
	\end{subfigure}
	\caption{The ice sheet, discretized by a mesh of tetrahedra. Color indicates the height of the base of the ice sheet (i.e., the mountain topography). The radius of the domain is $10^4$ meters, the maximum height of the mountain is $2.1 \times 10^3$ meters, and the average thickness of the ice sheet is 250 meters.}
	\label{fig:ice_mountain_mesh}
\end{figure}
